Constructing a universe for the setoid model

### 2020-01-08 types20-abstract.tex
small fixes

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be obtained as an inductive
recursive type.
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

However, we don't want to assume inductive-recursive types in the basic type
theory which is the target of the setoid model. We know that we can translate
basic instances of induction-recursion into inductive families using the
equivalence of families and discrete fibrations. For example, the
inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
\Set$ with $\Pi$-types and booleans is
\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
  this seems to be a non-conservative extension.}:
%
\begin{align*}
  & \textsf{data}\ \U\ : \Set_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
types in $\Set$ that are in the image of $\El$:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
\begin{align*}
  \U & = \Sigma (A : \Set) (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to an inductive-recursive universe in
$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
universe. Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop) (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

>>>>

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

However, we don't want to assume inductive-recursive types in the basic type
theory which is the target of the setoid translation. We know that we can translate
basic instances of induction-recursion into inductive families using the
equivalence of families and discrete fibrations. For example, the
inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
\Set$ with $\Pi$-types and booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
  this seems to be a non-conservative extension.}:
%
\begin{align*}
  & \textsf{data}\ \U\ : \Set_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
types in $\Set$ that are in the image of $\El$:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in
$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
universe. Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory.
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}
